---
title: "Introduction to canis, a `.can` file toolkit for R"
author: "Christophe Gigot"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output:
  rmarkdown::html_vignette:
    number_sections: yes
#    fig_width: 7
#    fig_height: 6
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Import a `.can` file

```{r setup, include=FALSE}
library(knitr)
library(rgl)
knit_hooks$set(rgl = hook_rgl)

# For the figures
myWidth <- 7
myEval <- TRUE
```

Let's start with importing a random `.can` file named `canopy.can`, and taking a look at the output.

```{r, include=FALSE}
library(canis)
#columbus <- readCan(system.file("examples/field.can", package="canis")[1])
plants <- readCan("~/Documents/can-files/examples/field.can") # pas propre du tout ici, recorus `a syseme et co
```

```r
library(canis)
plants <- readCan("/path/to/canopy.can")
```

```{r}
head(plants, 3)
str(plants)
```

We see that the first `r ncol(plants) - 1` columns correspond to different IDs or labels. Contrary to the other ones, the last column (`vertices`) is actually a so-called list-column. A key advantage is a structural independance from the number of vertices each polygon is made of. The package `canis` could therefore be extented to deal with many sorts of polygons (and not only triangles) without the need to totally redesign the interface.

If we glance at an element of the column `vertices`, we realize that it is a matrix. Each row corresponds to the coordinates of a vertex of the polygon (in a Cartesian coordinate system). The column names (`x`, `y` and `z`) refer to the three spatial dimensions.

```{r}
is(plants$vertices[[36]])
plants$vertices[[36]]
```

## Display in 3D

`plants`, the output of the function `readCan`, can be directly displayed in 3D using `display3d`. The `color` parameter may refer to a color (a color name, an output of the `rgb` function,...) or the name of a column of `plants` (except `vertices`). Type `?canis::display3d` for more information.

```{r, include=FALSE, eval=myEval}
view3d( theta = 0, phi = -75, zoom = 0.65)
```

```{r, rgl=TRUE, fig.width=myWidth, eval=myEval}
display3d(plants, col = "green4", alpha = 0.3)
```

```{r, include=FALSE, eval=myEval}
view3d( theta = 0, phi = -75, zoom = 0.65)
```

```{r, rgl=TRUE, fig.width=myWidth, eval=myEval}
display3d(plants, col = "plantID")
```

```{r, include=FALSE}
```

## Split the canopy in accordance with a set of volumes

### Design the set of volumes to play with

We need to create a set of volumes whose all the sides will define the boundaries to chop the canopy accordingly. To do so, we use the dedicated function `makeVolumeSet` which allows to build a set of volumes based on a regular grid.

```{r}
rangeDim <- function(polygon, dim) {
    range(sapply(polygon$vertices, function(i) i[, dim]))
}
x <- rangeDim(plants, "x")
y <- rangeDim(plants, "y")
z <- rangeDim(plants, "z")
volumes <- makeVolumeSet(x, y, z, intervals = c(3, 1, 3), label = "auto")
volumes
str(volumes)
```

Let's display this set of volumes above the canopy:

```{r, include=FALSE, eval=myEval}
view3d( theta = 0, phi = -75, zoom = 0.65)
```

```{r, rgl=TRUE, fig.width=myWidth, eval=myEval}
display3d(plants, col = "green4")
display3d(volumes, col = "volumeID", alpha = 0.3, label = "levels")
```

```{r, include=FALSE}
```

### Allocate a volume to each polygon (and shred polygons if needed)

Once in possession of these volumes, we can allocate each polygon of the scene to a volume using the function `splitCanopy`. To shred each polygon crossing a side of a volume, the parameter `shred` must be set to `TRUE` (default).

**Note:** In its current implementation, `canis` does not take into account overlapping of volumes. This means that if, for a given polygon, the first volume handled by the algorithm contains the polygon, it's done: `canis` will not check if this polygon is also included in another volume. So be careful, and remember: Visualize, visualize and visualize again your scenes before going further ahead!

```r
newPlants <- splitCanopy(plants, volumes, shred = TRUE,
                         parallel = TRUE, nCores = 8)
```

```{r, echo=FALSE}
warning(paste0("Parallelization is an experimental feature. ",
                           "In addition, progress bar may not be ",
                           "displayed when activated."))
```

```{r, include=FALSE}
filename <- ".RData"
if (file.exists(filename)) {
    load(filename)
} else {
    newPlants <- splitCanopy(plants, volumes, shred = TRUE,
                             parallel = TRUE, nCores = 8)
    save(newPlants, file = ".RData")
}
```

That's it. Let's check that we have a new column named `volumeID`.

```{r}
head(newPlants, 3)
str(newPlants)
```


```{r, include=FALSE, eval=myEval}
view3d( theta = 0, phi = -75, zoom = 0.65)
```

```{r, rgl=TRUE, fig.width=myWidth, eval=myEval}
display3d(newPlants, col = "volumeID")
display3d(volumes, col = "grey", alpha = 0.1)
```

```{r, include=FALSE}
```

Note that depending on your computer and the scene complexity (i.e. the number of polygons and volumes), 3D display may take a few tens of seconds. 

## And now?

### Save the output of `splitCanopy`

You may want to save the data frame returned by `splitCanopy` in a text file (e.g. a `.csv` file) for backup reasons or to use it in another software. To do so, you first need to flatten the data frame in order to get rid of the list-column structure. The function `flatten` does this job and return a regular data frame that you can seamlessly save in a text file.

```{r}
flatNewPlants <- flatten(newPlants)
filename <- tempfile(fileext = ".csv")
write.csv(flatNewPlants, filename, row.names = FALSE)
```

```{r, include=FALSE}
file.remove(filename)
```

### Perform analysis stuff in R

As we are already in a nice and rich working environment, let's try to plot a couple of graphes of our data set. We first need to import a few helpful R packages and write some relevant functions.

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

## A set of useful functions
# Area of a triangle computed using Heron's formula
heron <- function(triangle) {
    distances <- as.vector(dist(triangle))
    s <- sum(distances) / 2
    m <- s * (s - distances[1]) * (s - distances[2]) * (s - distances[3])
    # Even if it's not supposed to occur in theory, we need to check that m > 0.
    # Due to numeric limit, we can get negative numbers.
    if (m < 0) return(0)
    else       return(sqrt(m))
}
# Sum of all the areas of a list of triangles
area <- function(triangles) sapply(triangles, function(x) heron(x))
```

Now, let's go.

```{r}
myData <- newPlants %>%
    filter(leafID != 0) %>% # We get rid of the internodes
    mutate(area = area(vertices))
mySubData1 <- myData %>%
    group_by(plantID, leafID, volumeID) %>%
    summarise(leafAreaInVol = sum(area))
mySubData2 <- myData %>%
    group_by(plantID, leafID) %>%
    summarise(leafArea = sum(area))
myData <- left_join(mySubData1, mySubData2, by = c("plantID", "leafID")) %>%
    mutate(pcentLeafAreaInVol = leafAreaInVol / leafArea) %>%
    select(-leafArea) %>%
    separate(volumeID, into = c("x", "y", "z"), sep = "-") %>%
    as.data.frame()
rm(mySubData1, mySubData2)
head(myData)
```

```{r, fig.width=myWidth}
# Leaf area density in a 2D x-z space
myDataA <- myData %>%
    group_by(x, z) %>%
    summarise(leafAreaInVol = sum(leafAreaInVol))
myDataA$x <- as.factor(myDataA$x); levels(myDataA$x) <- c("left", "middle", "right")
myDataA$z <- as.factor(myDataA$z); levels(myDataA$z) <- c("low", "middle", "high")
g <- ggplot(myDataA, aes(x, z)) +
    geom_raster(aes(fill = leafAreaInVol)) +
    scale_fill_gradient(low = "white", high = "green4",
                        name = expression("Leaf area ("~cm^2~")")) +
    theme_classic() +
    labs(x = "Canopy side", y = "Canopy height",
         title = "Leaf area density in a 2D x-z space")
print(g)
```

```{r, fig.width=myWidth}
# Leaf area density in a 2D "leaf level-z space"
test6 <- myData %>%
    group_by(leafID, z) %>%
    summarise(leafAreaInVol = sum(leafAreaInVol))
test6$z <- as.factor(test6$z); levels(test6$z) <- c("low", "middle", "high")
g <- ggplot(test6, aes(leafID, z)) +
    geom_raster(aes(fill = leafAreaInVol)) +
    scale_fill_gradient(low = "white", high = "green4",
                        name = expression("Leaf area ("~cm^2~")")) +
    theme_classic() +
    labs(x = "Leaf level", y = "Canopy height",
         title = "Leaf area density in a 2D \"leaf level-z space\"")
print(g)

```

```{r, include=FALSE}
```


